Visualizing Contingency Tables

### 2006-04-10 trunk/vcd/paper/hsc.Rnw
went through HSC revision
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@518 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%% Hyperref
\hypersetup{%
  pdftitle = {Visualizing Multi-way Contingency Tables},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {David Meyer, Achim Zeileis, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

<<preliminaries,echo=FALSE,results=hide>>=
myseed <- 4711
set.seed(myseed)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")

set.seed(myseed)
#art_max <- coindep_test(art, n = 5000)
uc_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & SAS & S-PLUS & R & ViSta & MANET/Mondrian\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} (see Tab.~\ref{tab:hospital}).

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable in (hypothesized) causal relationships.

@@

@@

In the last section, we have seen how to compare observed and expected
values of a contingency table using sieve plots. We can do this
more straightforwardly by using a plot that directly visualizes the residuals. The
most widely known residuals are the Pearson residuals

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If causal
relationships between the variables are assumed, the determining variables
should be used first for splitting since they will dominate the
plot. Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. However, for nominal variables, mosaic plots are
a better choice, especially when residual-based 
shadings are used (see next section), Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary concern.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
\label{sec:colors}

@@

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e. appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

For association plots, residual-based shadings are redundant since all relevant
information is already contained in the plot by construction. But using one of the
shadings discussed above will nevertheless 
support the analysis process and is therefore recommended.
For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
on the one hand facilitates the discrimination between
positive and negative residuals, and on the other, more importantly,
allows an easier comparison of the tiles' sizes. Thus, the shading
supports the detection of the pattern.

@@

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third piece of information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using darker (`uninteresting') colors for non-significant
results. These can again be derived using the interpolation procedure
described in Sect.~\ref{sec:colors} by
restricting the luminance to a shorter range, ending with a ``darker'' value.

% The following example using the \data{Bundesliga} data \cite{vcd:Knorr-Held:1999}
% shows the relationship of home goals and away
% goals of Germany's premier soccer league in 1995: although there are two
% ``larger'' residuals (one greater than 2, one less then $-2$), the
% $\chi^2$ test does not reject the null hypothesis of
% independence on a significance level of $95\%$ since the $p$ value is $0.1213$. 
% Consequently, the employed shading uses a darker overall color scheme
% (see Fig.~\ref{fig:bundesliga}).

@@

\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $99\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Fig.~\ref{fig:arthritis2}, we again
visualize the \data{arthritis} data, this
time using the maximum test statistic. The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@
\caption{Mosaic plot for the \data{Arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, gp = shading_max)
@
\caption{Mosaic plot for the \data{Arthritis} data, using the maximum test and 
data-driven cut-off points for the residuals.}
\label{fig:arthritis2}
\end{center}
\end{figure}

@@

Especially for the visualization of areas, 
use perceptually uniform colors and color palettes, derived from,
e.g., the HCL color space. Shadings can be used to add 
information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis of causal relationships.
Residual-based shadings can be used for model diagnostics, e.g.,
by using the hue to code the sign of a residual, and a linear
combination of chroma and luminance for the size. In addition, the
significance of test statistics can be visualized through overall 
darker color palettes. Using the maximum instead of the $\chi^2$ test
statistic to aggregate the residuals allows precise diagnostics when
the null hypothesis of independence is rejected.

@@

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and also varies given department.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
pairs(UCBAdmissions)
@
\caption{Pairs plot for the \data{HairEyeColor} data.}
\label{fig:pairs}
\end{center}
\end{figure}

>>>>

%% Hyperref
\hypersetup{%
  pdftitle = {Visualizing Contingency Tables},
  pdfsubject = {Handbook of Statistical Computing},
  pdfkeywords = {association plots, contingency tables, HCL colors, HSV colors, mosaic plots, sieve plots},
  pdfauthor = {David Meyer, Achim Zeileis, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

<<preliminaries,echo=FALSE,results=hide>>=
myseed <- 1071
set.seed(myseed)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")

set.seed(myseed)
art_max <- coindep_test(art, n = 5000)

set.seed(myseed)
ucb_max <- coindep_test(aperm(UCBAdmissions, c(3,2,1)), margin = "Dept", n = 5000)

@@

\begin{itemize}
  \item Varying the shape of mosaic plots (as well as bar plots) to yield, e.g.,
  double-decker plots \cite{vcd:Hofmann:2001}, spine 
  plots \cite{vcd:riedwyl+schuepbach:1994}, or spinograms \cite{vcd:hofmann+theus}.
\item Using residual-based shadings to visualize log-linear models
  \cite{vcd:Friendly:1994,vcd:Friendly:2000} and significance of
  statistical tests \cite{vcd:Meyer+Zeileis+Hornik:2003}.
\item Using pairs plots and trellis-like layouts for marginal, conditional and
  partial views \cite{vcd:Friendly:1999}.
\item Adding direct user interaction, allowing quick exploration and
  modification of the visualized models \cite{vcd:Unwin+Hawkins+Hofmann:1996,vcd:Theus:2003}.
\item Providing a modular and flexible implementation to easily allow
  user extensions \cite{vcd:Meyer+Zeileis+Hornik:2005a}.
\end{itemize}

@@

\begin{table}[h]
  \centering
  \begin{tabular}{|l|c|c|c|c|c|}
    \hline
                             & \proglang{SAS} & \proglang{S-PLUS} & \proglang{R} & \pkg{ViSta} & \pkg{MANET}/\pkg{Mondrian}\\\hline
    Basic functionality      &  $\times$  &   $\times$    & $\times$ &   $\times$   & $\times$\\
    Shape                    &     &        & $\times$ &       & $\times$\\
    Residual-based shadings  &  $\times$  &        & $\times$ &   $\times$   & ($\times$)\\
    Conditional Views        &  $\times$  &        & $\times$ &       & $\times$\\
    Interaction              &     &        &   &   $\times$   & $\times$\\
    Extensible Design        &     &        & $\times$ &       &  \\\hline
  \end{tabular}
  \caption{Comparison of current software environments.}
  \label{tab:compare}
\end{table}
%FIXME% For the JSS paper there was criticism about haven MANET and Mondrian in the
%FIXME% same column, how did we decide ro resolve this? Different columns, I suppose...?!

@@

Throughout this section, our examples will be based on
the \data{hospital} data \cite{vcd:wing:1962} given in Tab.~\ref{tab:hospital}.

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
it seems indeed that LOS is homogeneous with respect to these two visit frequency strata.

@@

Since mosaic plots are asymmetric by construction, the choice of the
variable order matters, as the first splitting variable dominates the 
plot. In our example, if we use `Visit frequency' %'
as the first splitting variable, the impression is very different compared to the previous
mosaic (see Fig.~\ref{fig:mosaic3fig}). 
In this alternative display, we see the marginal distribution of `Visit frequency' %'
in the rows: about half of the patients get visited regularly. This
group is dominated by patients staying between 2 and 9 years.
It seems apparent that the distribution of LOS %'
is similar for monthly and never visited patients, so this two categories actually
represent one homogenous group (patients visited only casually). Since
the first splitting variable dominates the plot, it should be chosen
as to be the explanatory variable.
%FIXME% causality is not needed, hence omitted: in (hypothesized) causal relationships.

@@

%FIXME% Maybe formulate the following theory a bit more like:
%FIXME% "To fix notations, ...", "It is well-known that..."

@@

In the last section, we have seen how to compare observed and expected
values of a contingency table using sieve plots. We can do this
more straightforwardly by using a plot that directly visualizes the residuals. The
most widely used residuals are the Pearson residuals

@@

Mosaic plots are an instrument to visualize the \emph{observed} frequencies
of a contingency table by recursive conditional splits. If the variables can be
separated into explanatory and dependent variable(s), the explanatory variable(s)
should be be used first for splitting to display the conditional distribution of
the dependent variable(s) given the explanatory variable(s). %FIXME% causality removed again
Sieve plots basically visualize the table of \emph{expected} frequencies, and
in addition the deviations from the observed frequencies by the density of a
grid added to each tile. They complement mosaic plots in detecting dependency
patterns for ordinal variables. An alternative way of enhancing mosaic plots
to display deviations from expected frequencies is to use
residual-based shadings (see next section) which are typically more intellegible
than sieve plots, in particular for nominal variables.
Association plots directly
visualize Pearson and raw residuals, i.e., standardized and
non-standardized deviations of observed from expected frequencies,
respectively. These plots should be used if the diagnostics of
independence models is of primary interest.

@@

\subsection[A Note on Colors and Color Palettes]{A Note on Colors and Color Palettes\footnote{The electronic version of this article uses colors.}}
\label{sec:colors} 
%FIXME% Maybe extend the footnote and say that the printed version is monochrome,
%FIXME% and that the electronic version uses color.
%FIXME% We can already say earlier that the R code for reproducing the figures
%FIXME% is available from a Web page. And then we can say here that the colored versions
%FIXME% can also be obtained from the Web page.
%FIXME% Where do we want to set up the URL? I would suggest
%FIXME%   http://statmath.wu-wien.ac.at/projects/vcd/

@@

Choosing colors and color palettes in RGB space can be a bit inconvenient
and hence software implementations are often based on
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Satu\-ration 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in HSV space is
represented by three dimensions $(H, S, V)$: the hue $H$ (dominant wavelength in the spectrum, in $[0, 360]$), 
the saturation $S$ (`colorfulness' or `pureness', in $[0, 100]$), and 
the value $V$ (`brightness', amount of gray, in $[0, 100]$).\footnote{In many implementations,
all three dimensions are scaled to the unit interval instead of the coordinates used here.}
These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
However, HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette (colors $(H, 100, 100)$ for varying hues $H$) in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red $(0, 100, 100)$
over a neutral white $(H, 0, 100)$ to a saturated blue $(240, 100, 100)$.
Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<colorfig,fig=TRUE,echo=FALSE,height=3,width=7,results=hide>>=
par(mfrow = c(1,2), mar = c(1,1,1,1), oma = c(0,0,0,0))
pie(rep(1,9), radius = 1, col = rainbow(9), labels = 360 * 0:8/9)
pie(rep(1,9), radius = 1, col = rainbow_hcl(9), labels = 360 * 0:8/9)
@ 
\caption{Qualitative color palette for the HSV (left) 
         and HCL (right) spaces. The HSV colors are $(H, 100, 100)$ and
	 the HCL colors $(H, 50, 70)$ for the same hues $H$. Note that
         in a monochrome version of this paper, all pies in the right
         wheel will be shaded with the same gray, i.e., appear identical.}
\label{fig:colors}
\end{center}
\end{figure}

@@

The use of colors that are more `in harmony' goes back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
These colors are again specified by triplets $(H, C, L)$:
chroma $C$ loosely corresponds to colorfulness and luminance $L$ to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and using varying hues. HCL colors with fixed luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging HCL color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} by interpolating again between a
neutral color such as $(H, 0, 90)$ and two colors with full chroma
such as blue $(260, 100, 50)$ and red $(0, 100, 50)$. The resulting
palette is shown in the lower part of Fig.~\ref{fig:diverging}. 
In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus receive the same perceptual ``weight''.
Note, that chroma and luminance are varied simultaneously, i.e.,
the full chroma colors are also darker than the neutral color. Of course,
it would also be possible to choose a darker neutral color (or lighter 
full chroma colors), however, by varying both chroma and luminace better 
contrasts can be achieved.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

In Fig.~\ref{fig:Friendly}, we show again the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).
\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}
Clearly, the asymmetry for regular and never visited students,
and the pattern inversion for lengths of stay of 2--9 and more than 19
years are emphasized using the color shading.

For association plots, residual-based shadings are redundant since all relevant
information is already contained in the plot by construction. Nevertheless, using one of the
shadings discussed above will support the analysis process and is therefore recommended.
For example, applying the Friendly shading in Fig.~\ref{fig:colorassoc} 
on the one hand facilitates the discrimination between
positive and negative residuals, and on the other, more importantly,
allows an easier comparison of the tiles' sizes. Thus, the shading
supports the detection of the pattern.

@@

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third piece of information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using less colorful (`uninteresting') colors for non-significant
results. These can again be derived using the same procedure
described in Sect.~\ref{sec:colors} but using a smaller amount of color, i.e.,
a smaller maximal chroma (e.g., 20 instead of 100).

% The following example using the \data{Bundesliga} data \cite{vcd:Knorr-Held:1999}
% shows the relationship of home goals and away
% goals of Germany's premier soccer league in 1995: although there are two
% ``larger'' residuals (one greater than 2, one less then $-2$), the
% $\chi^2$ test does not reject the null hypothesis of
% independence on a significance level of $5\%$ since the $p$ value is $0.1213$. 
% Consequently, the employed shading uses a darker overall color scheme
% (see Fig.~\ref{fig:bundesliga}).

@@

\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $1\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Fig.~\ref{fig:arthritis2}, we visualize again the \data{arthritis} data, this
time using the maximum test statistic and its 10\% and 1% critical values at cut-off points.
The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@
\caption{Mosaic plot for the \data{arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis2fig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
set.seed(myseed)
mosaic(art, gp = shading_max, gp_args = list(n = 5000))
@
\caption{Mosaic plot for the \data{arthritis} data, using the maximum test and 
data-driven cut-off points for the residuals.}
\label{fig:arthritis2}
\end{center}
\end{figure}

@@

For the visualization of areas perceptually uniform colors and color palettes
should always be used and HCL color space provides a convenient way to do so.
Shadings can be used to add information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis relationships with a 
single dependent variable. Residual-based shadings can be used for model diagnostics, e.g.,
by using diverging color palettes to code the sign and size of a residual.
In addition, the significance of test statistics can be visualized by using 
overall less colorful palettes. Using the maximum instead of the $\chi^2$ (or other) test
statistic(s) allows for data-driven cut-offs in the diverging palettes and
precise diagnostic identification of the cells in conflict with the
null hypothesis of independence.

@@

A first step in the explorative analysis of more complex tables is to
get a quick overview on the data. For that, 
all basic plots can be combined in pairwise displays,
arranged in in a matrix similar to scatterplots in a pairs plot.
The diagonal cells contain the variable names, optionally with
univariate statistics, whereas the off-diagonal cells contain plots whose
variables are implicitly specified by the cells' position in the matrix. %'
In Fig.~\ref{fig:pairs}, the diagonal cells show bar plots for the
distribution of the variables, and the off-diagonal cells mosaic plots
for the corresponding pair of variables. The plots suggest that
admission differs between male and female students, and between the departments,
and that the proportion of male and female students varies across the departments.

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<pairs,fig=TRUE,echo=FALSE,height=8,width=8,results=hide>>=
pairs(UCBAdmissions)
@
\caption{Pairs plot for the \data{UCB admissions} data.}
\label{fig:pairs}
\end{center}
\end{figure}
